---
title: "Sanity Report"
author: Kyle Kimler
date: June 28, 2020
output:
  html_document:
    keep_md: true
    toc: true
    number_sections: false
    toc_float: true
---

Introduction
---------------

Just testing out Seurat using the nasal polyp dataset from Jose's 2018 paper
Allergic inflammatory response


Chapter 1: Loading data to Seurat
---------------

Load libraries
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(Seurat)
```

```{r, warning=FALSE}
umifile <- read.csv(file="500sampPandasGenePolyp.txt",header=TRUE,stringsAsFactors=FALSE)
umidata <- umifile %>% remove_rownames %>% column_to_rownames(colnames(umifile)[1])

#grep rownames to find genes of interest - in this example opsins
umidata %>% filter(grepl("OPN1",row.names(umidata))) %>% rownames

polyp <- CreateSeuratObject(umidata, project = "polyp_scRNAseq")

#look at specific genes in the seurat object
polyp[c("OPN1LW", "ROPN1", "ROPN1L"), 1:30]

#use [[]] puts a new metadata column to the seurat object
polyp[["percent.mt"]] <- PercentageFeatureSet(polyp, pattern = "^MT-")

#look at metadata summaries of seurat object
head(polyp@meta.data, 5)

```

```{r}
#print violin plots of mitochondrial genes
VlnPlot(polyp, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

#more QC comparisons
plot1 <- FeatureScatter(polyp, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(polyp, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

Chapter 2: Normalization
---------------
First, I'm following the seurat vignette and just using log transform normalization. I want to try Sanity later.

```{r}
polyp <- NormalizeData(polyp, normalization.method = "LogNormalize", scale.factor = 10000)
```

Next, we pick out the most variable genes from the dataset.
```{r, warning=FALSE}

polyp <- FindVariableFeatures(polyp, selection.method = "vst", nfeatures = 2000)

#what are the top 10?
top10 <- head(VariableFeatures(polyp), 10)

#and plot those highly variable genes with and without labels
plot1 <- VariableFeaturePlot(polyp)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)

#remove legend from first plot and y-axis shit from second
plot1 + theme(legend.position="none") + plot2 + theme(axis.text.y=element_blank())+ylab("") + scale_y_discrete(breaks=NULL)

```

Chapter 3: Data scaling and visualizing dimensionality reduction
---------------

Seurat uses a "ScaleData" function:
  - shift the expression of each gene so that mean expression across cells is 0
  - scale the expression of each gene, so that the variance across cells is 1

```{r}
all.genes <- rownames(polyp)
polyp <- ScaleData(polyp, features = all.genes)

```

And next, let's perform PCA
```{r}
polyp <- RunPCA(polyp, features = VariableFeatures(object = polyp))

```

We can visualize PCA in several different ways

First, let's look at the highest contributing genes to each component
```{r}
VizDimLoadings(polyp, dims = 1:5, reduction = "pca")
```

And we can look at the distribution of cells by the components we like
```{r}
DimPlot(polyp, reduction = "pca")
```

Or look at heatmaps for the biggest gene contributors across cells
```{r}
DimHeatmap(polyp, dims = 1, cells = 499, balanced = TRUE)
```

For any number of components
```{r}
DimHeatmap(polyp, dims = 1:9, cells = 499, balanced = TRUE)
```

Chapter 4: How many useful dimensions are in the dataset?
-------------
Seurat uses Macosko et al 2015 method of choosing number of important dimensions,
where PCA is run many times on 1% of the dataset, like bootstrapping.
```{r}
#run the bootstrap
polyp <- JackStraw(polyp, num.replicate = 100)
#score the resulting dimensions
polyp <- ScoreJackStraw(polyp, dims = 1:20)

#and plot distribution of p-values for each component
JackStrawPlot(polyp, dims = 1:20)

#also create an elbow plot of the variance explained by each component
ElbowPlot(polyp)
```

Hard to tell whether to cut off at 5 or at 13 so I'll try both.

Chapter 5: Clustering the cells with UMAP and t-sne
-------------

```{r}
polyp <- FindNeighbors(polyp, dims = 1:5)
polyp <- FindClusters(polyp, resolution = 0.5)
head(Idents(polyp), 5)

```

Plot~!
```{r}
polyp <- RunUMAP(polyp, dims = 1:5)
DimPlot(polyp, reduction = "umap")
```

Looks bad with 5. Now with 13 components


```{r}
polyp <- FindNeighbors(polyp, dims = 1:13)
polyp <- FindClusters(polyp, resolution = 0.5)
head(Idents(polyp), 13)

```

Plot~!
```{r}
polyp <- RunUMAP(polyp, dims = 1:13)
DimPlot(polyp, reduction = "umap")

#to point out where your favorite genes are living
FeaturePlot(polyp, features = c("OPN1LW", "ROPN1", "ROPN1L"))
```

Chapter 6: Finding biomarkers with Seurat
------------

Find markers for all of the cell types displayed in the UMAP
```{r}

polyp.markers <- FindAllMarkers(polyp, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

#print the best markers
polyp.markers %>% group_by(cluster) %>% top_n(n = 2, wt = avg_logFC)

#put the gene names in a list
polyp.markers %>% group_by(cluster) %>% top_n(n = 2, wt = avg_logFC) -> test
biomarkers <- unlist(as.list(test$gene))
```

```{r, warning=FALSE}
#Display where those genes appear in the cell types of the UMAP
FeaturePlot(polyp, features = biomarkers, pt.size=0.5)
```

Finally, display the presence of our biomarkers in cells of each cell type by heatmap
```{r}
top10 <- polyp.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_logFC)
DoHeatmap(polyp, features = top10$gene) + NoLegend()
```


```{r}
DimPlot(polyp, reduction = "umap")

```

To save your end plot, run this:

saveRDS(pbmc, file = "../output/pbmc_tutorial.rds")


Till next time~
